A High-Order Discontinuous Galerkin Solver with Dynamic Adaptive Mesh Refinement to Simulate Cloud Formation Processes

### 2019-10-29 writing/paper/paper.tex
Add some improvements to paper.

\begin{document}
\maketitle 
\begin{abstract}
  We present a high-order discontinuous Galerkin (\dg{}) solver of the compressible Navier-Stokes equations for cloud formation processes.
  The scheme exploits an underlying parallelized implementation of the \aderdg{} method with dynamic adaptive mesh refinement. 
  We improve our method by a \pde{}-independent general refinement criterion, based on the local total variation of the numerical solution.
  Our generic scheme shows competitive results for both classical \textsc{cfd} and stratified scenarios.
  While established methods use numerics tailored towards the specific simulation, our scheme works scenario independent.
  All together, our method can be seen as a perfect candidate for large scale cloud simulation runs on current and future super computers.

@@

\section{Results}
We use a mix of various benchmarking scenarios that consist of standard \textsc{cfd} scenarios and atmospheric flows.
%
For the standard \textsc{cfd} testing scenarios, we use the following constants of fluids:
\begin{equation}
  \gamma = 1.4, \quad \Pr = 0.7, \quad c_v = 1.0.
\end{equation}
A simple scenario is the Taylor-Green vortex.
It has an analytical solution in the incompressible limit which is given by
\begin{align}
  \label{eq:taylor-green}
  \begin{split}
  \Qrho(\bm{x}, t) &= 1,\\
  \Qv(\bm{x}, t) &= \exp(-2 \mu t)
  \begin{pmatrix}
    \phantom{-}\sin(x) \cos(y) \\
- \cos(x) \sin(y) 
    \end{pmatrix}, \\
  \pressure(\bm{x}, t) &= \exp(-4 \mu t) \, \nicefrac{1}{4} \left( \cos(2x) + \cos(2y) \right) + C.
  \end{split}
\end{align}
The constant $C = \nicefrac{100}{\gamma}$ governs the speed of sound and thus the Mach number $\text{Ma} = 0.1$~\cite{dumbser2016high}. The viscosity is set to $\mu = 0.1$.

We simulate on the domain $[0,2\pi]^2$ 
%We use a domain of size 
%$(\SI{2 \pi}{\m} \times \SI{2 \pi}{\m})$ 
%% keine Meter-Angaben bei analytischen Testcases
and impose the analytical solution at the boundary.
A comparison at time $t = 10.0$ of the analytical solution for the pressure with our approximation (\cref{fig:taylor-green}) shows excellent agreement.
We used an \aderdg{}-scheme of order $5$ with a grid of $25^2$ cells.

@@

The Arnold-Beltrami-Childress (\textsc{abc}) flow is similar to the Taylor-Green vortex but possesses an analytical solution for three-dimensions as well~\cite{tavelli2016staggered}.
It is defined in the domain \( \left[ -\pi, \pi \right]^3 \) as
\begin{align}
  \label{eq:abc-flow}
  \begin{split}
  \Qrho(\bm{x}, t) &= 1,\\
  \Qv(\bm{x}, t) &= \phantom{-} \exp(-1\mu t)
  \begin{pmatrix}
    \sin(z) + \cos(y)\\
    \sin(x) + \cos(z)\\
    \sin(y) + \cos(x)
  \end{pmatrix}, \\
  \pressure(\bm{x}, t) &= -\exp(-2 \mu t) \, \left(\cos(x)\sin(y) + \sin(x)\cos(z) + \sin(z)\cos(y)\right)
  + C.
  \end{split}
\end{align}
again under the assumption of incompressibility.
The constant $C = \nicefrac{100}{\gamma}$ is chosen as before.
We use a viscosity of $\mu = 0.01$ and analytical boundary conditions.
Our results (\cref{fig:abc-flow}) show a good agreement between the analytical solution and our approximation with an \aderdg{}-scheme of order $3$ with a mesh consisting of $27^3$ cells at time $t = \SI{0.1}{\s}$.

@@

With  the constants
\begin{equation}\label{eq:atmosphere-constants}
    \gamma = 1.4 ,\quad \Pr =  0.71 ,\quad R = 287.058 ,\quad p_0 = \SI{10e5}{\Pa}, \quad g = \SI{9.8}{m/s^2},
\end{equation}
all following stratified flow scenarios are described in terms of the potential temperature
\begin{equation}
  \potT = T \left( \frac{p_0}{p} \right)^{R/c_p},
\end{equation}
with reference pressure $p_0$~\cite{muller2010adaptive,giraldo2008study}.
%
We compute the initial background density and pressure by inserting the assumption of a constant background energy in \cref{eq:hydrostatic-balance}.
The background atmosphere is then perturbed.
We set the density and energy at the boundary such that it corresponds to the background atmosphere.
Furthermore, to ensure that the atmosphere stays in hydrostatic balance, we need to impose the viscous heat flux
\begin{equation}
  \label{eq:atmosphere-bc}
  \viscFlux_{\QE} = \kappa \pdv{\overline{T}}{z}.% =
\end{equation}
at the boundary~\cite{giraldo2008study}.
In this equation, $\overline{T}(z)$ is the background temperature at position $z$, which can be computed from \cref{eq:hydrostatic-balance,eq:eos}.

@@

\todo{Mention simulation order for two bubbles!}
We have two bubbles, with constants
\begin{equation}
  \label{eq:bubbles-values}
\begin{alignedat}{6}
  & \text{warm:} \qquad && A = \SI{0.5}{\K}, \quad&& a = \SI{150}{\m}, \quad&& s = \SI{50}{\m}, \quad&& x_c = \SI{500}{\m,} \quad&& z_c = \SI{300}{\m},\\
  & \text{cold:} \qquad && A = \SI{-0.15}{\K}, \quad&& a = \SI{0}{\m}, \quad&& s = \SI{50}{\m}, \quad&& x_c = \SI{560}{\m}, \quad&& z_c = \SI{640}{\m}.
  \end{alignedat}
\end{equation}
Similar to~\cite{muller2010adaptive}, we use a constant viscosity of $\mu = 0.01$ to regularize the solution.
Note that we use a different implementation of viscosity than~\cite{muller2010adaptive}.
Hence, it is difficult to compare the parametrization directly.
We ran this scenario twice: once without \amr{} and a mesh of size $\SI{1000/81}{\m} = \SI{12.35}{\m}$ and 
once with \amr{} with two adaptive refinement levels and parameters $\Trefine = 2.5$ and $\Tdelete = -0.5$.
We specialize the \amr{}-criterion~\pcref{eq:refinement-criterion} to our stratified flows by using the potential temperature.
This resulted in a mesh with cell-size lengths of approx.\ \SI{111.1}{\m}, \SI{37.04}{\m}, and \SI{12.34}{\m}.
The resulting mesh can be seen in \cref{fig:two-bubbles-ader}.
We observe that the $L_2$ difference between the potential temperature of the \amr{} run, which uses 1953 cells, and the one of the fully refined run with 6561 cells, is only $1.87$.
The relative error is \num{6.69e-6}.
We further emphasize that our \amr{}-criterion accurately tracks the position of the edges of the cloud instead of only its position.
This is the main advantage of our gradient-based method in contrast to methods working directly with the value of the solution, as for example~\cite{muller2010adaptive}.
Overall, our result for this benchmark shows an excellent agreement to the previous solutions of~\cite{muller2010adaptive}.

@@

For our second scenario, the cosine bubble, we use a perturbation of the form
\begin{align}
  \label{eq:cos-pertubation}
  \pertubationPotT &= \begin{cases}
    \nicefrac{A}{2} \left[ 1 + \cos(\pi r) \right] & r \leq a, \\
    0 & r > a,
    \end{cases}
\end{align}
where $A$ denotes the maximal perturbation and $a$ is the size of the bubble.
We use the constants
\begin{equation}\label{eq:cosine-bubble}
  A = \SI{0.5}{\K} \quad a = \SI{250}{\m} \quad x_c = \SI{500}{\m} \quad z_c = \SI{350}{\m}.
\end{equation}
For the three-dimensional bubble, we set $y_c = x_c = \SI{500}{\m}$.
This corresponds to the parameters used in~\cite{kelly2012continuous}\footnote{%
We found that the parameters presented in the manuscript of~\cite{kelly2012continuous} only agree with the results, if we use the same parameters as for 2D simulations.}.
For the 2D case, we use a constant viscosity of $\mu = 0.01$ and an \aderdg{}-method of order 5 with two levels of dynamic \amr{}, resulting again in cell sizes of roughly $\SI{111.1}{\m}, \SI{37.04}{\m}, \SI{12.34}{\m}$.
We use slightly different \amr{} parameters of $\Trefine = 1.5$ and $\Tdelete = -0.5$ and let the simulation run for \SI{600}{\s}.
Note that, as seen in \cref{fig:cosine-2d}, our \amr{}-criterion tracks the wavefront of the cloud accurately.
This result shows an excellent agreement to the ones achieved in~\cite{giraldo2008study,muller2010adaptive}.

@@

@@

>>>>

\begin{document}
\maketitle 
\begin{abstract}
  We present a high-order discontinuous Galerkin (\dg{}) solver of the compressible Navier-Stokes equations for cloud formation processes.
  The scheme exploits an underlying parallelized implementation of the \aderdg{} method with dynamic adaptive mesh refinement. 
  We improve our method by a \pde{}-independent general refinement criterion, based on the local total variation of the numerical solution.
  Our generic scheme shows competitive results for both classical \textsc{cfd} and stratified scenarios.
  While established methods use numerics tailored towards the specific simulation, our scheme works scenario independent.
  All together, our method can be seen as a perfect candidate for large scale cloud simulation runs on current and future super computers.\todo{Less bragging!}

@@

\section{Results}
In this section, we evaluate the quality of the results of our numerical methods and the scalability of our implementation.
We use a mix of various benchmarking scenarios.
After investigating the numerical convergence rate, we look at three standard \textsc{cfd} scenarios: the Taylor-Green vortex, the three-dimensional Arnold-Beltrami-Childress scenario and a lid-driven cavity flow.
Finally, we evaluate the performance for stratified flow scenarios both in two and three dimensions.

\subsection{CDF Testing Scenarios}
We begin with a manuactured solution scenario which we can use for a convergence test.
\todo{Introduce need for MMS}
Unless mentioned otherwise, we use the following constants of fluids:
\begin{equation}
  \gamma = 1.4, \quad \Pr = 0.7, \quad c_v = 1.0.
\end{equation}
Our description of the manufactured solution follows~\cite{dumbser2010arbitrary}.
To construct this solution, we assume that
% \begin{align}\label{eq:manufactured-solution}
% \begin{split}
%   \pressure(\bm{x}, t) &= \pressure_0 \cos( \bm{k} \bm{x} - \omega t ) + \pressure_b, \\
%   \Qrho(\bm{x}, t) &= \Qrho_0 \sin (\bm{k} \bm{x} - \omega t) + \Qrho_b, \\
%   \Qv(\bm{x}, t) &= \bm{v_0} \sin(\bm{k} \bm{x} - \omega t),
% \end{split}
% \end{align}
\begin{align}\label{eq:manufactured-solution}
  \begin{split}
    \pressure(\bm{x}, t) &= 0.1 \cos( \bm{k} \bm{x} - 2 \pi t ) + \gamma^{-1}, \\
    \Qrho(\bm{x}, t) &= 0.5 \sin (\bm{k} \bm{x} - 2 \pi t) + 1, \\
    \Qv(\bm{x}, t) &= \bm{v_0} \sin(\bm{k} \bm{x} - 2 \pi t),
  \end{split}
  \end{align}
solves our \pde{}.
We use the constants $\bm{v_0} = \nicefrac{1}{4} \left( 1, 1 \right)^\intercal$ and $\bm{k} = \nicefrac{\pi}{5} \left( 1, 1 \right)^\intercal$ 
% \begin{equation}
% \begin{alignedat}{6}
%   & \pressure_u &&= \nicefrac{1}{10}, \qquad& \pressure_b &= \nicefrac{1}{\gamma}, \\
%   & \Qrho_0 &&= \nicefrac{1}{2}, \qquad& \Qrho_b &= 1, \\
%   & \bm{v_0} &&= \nicefrac{1}{4},
%   \begin{pmatrix}
%     1 \\
%     1
%   \end{pmatrix},
%   \qquad& \bm{k} &= \nicefrac{\pi}{5}
%   \begin{pmatrix}
%     1\\
%     1
%   \end{pmatrix},\\
%   & \omega &&= 2 \pi.
%   \end{alignedat}
% \end{equation}
Note that \Cref{eq:manufactured-solution} does not solve the compressible Navier-Stokes equations \cref{eq:equation-set} directly.
It rather solves our equation set with an added source term which can be derived with a computer algebra system.
We ran this scenario for a grid of combinations of order $1, \ldots 6$ and mesh sizes of $??? $.
For this scenario, we achieve high-order convergence (\ref{fig:convergence-test}) but notice some diminishing returns for very large orders.
\begin{figure}[htb]
  \centering
  ??? 
  \caption{\label{fig:convergence-test}Mesh size vs.\ error for various orders.
  Dashed lines show the theoretical order of convergence.}
\end{figure}

After we have established that the implementation of our numerical method converges, we are going to investigate three established testing scenarios from the field of computational fluid mechanics.
A simple scenario is the Taylor-Green vortex.
Assuming an incompressible fluid, it can written as
\begin{align}
  \label{eq:taylor-green}
  \begin{split}
  \Qrho(\bm{x}, t) &= 1,\\
  \Qv(\bm{x}, t) &= \exp(-2 \mu t)
  \begin{pmatrix}
    \phantom{-}\sin(x) \cos(y) \\
- \cos(x) \sin(y) 
    \end{pmatrix}, \\
  \pressure(\bm{x}, t) &= \exp(-4 \mu t) \, \nicefrac{1}{4} \left( \cos(2x) + \cos(2y) \right) + C.
  \end{split}
\end{align}
The constant $C = \nicefrac{100}{\gamma}$ governs the speed of sound and thus the Mach number $\text{Ma} = 0.1$~\cite{dumbser2016high}. The viscosity is set to $\mu = 0.1$.

We simulate on the domain $[0,2\pi]^2$ and impose the analytical solution at the boundary.
A comparison at time $t = 10.0$ of the analytical solution for the pressure with our approximation (\cref{fig:taylor-green}) shows excellent agreement.
Note that we only show a qualitative analysis because this is not an exact solution for our equation set as we do assume compressibility of the fluid.
This is nevertheless a valid comparison because for very low Mach numbers, both incompressible and compressible equations behave in a very similar manner.
We used an \aderdg{}-scheme of order $5$ with a grid of $25^2$ cells.

@@

The Arnold-Beltrami-Childress (\textsc{abc}) flow is similar to the Taylor-Green vortex but is an analytical solution for the three-dimensions incompressible Navier-Stokes equations~\cite{tavelli2016staggered}.
It is defined in the domain \( \left[ -\pi, \pi \right]^3 \) as
\begin{align}
  \label{eq:abc-flow}
  \begin{split}
  \Qrho(\bm{x}, t) &= 1,\\
  \Qv(\bm{x}, t) &= \phantom{-} \exp(-1\mu t)
  \begin{pmatrix}
    \sin(z) + \cos(y)\\
    \sin(x) + \cos(z)\\
    \sin(y) + \cos(x)
  \end{pmatrix}, \\
  \pressure(\bm{x}, t) &= -\exp(-2 \mu t) \, \left(\cos(x)\sin(y) + \sin(x)\cos(z) + \sin(z)\cos(y)\right)
  + C.
  \end{split}
\end{align}
The constant $C = \nicefrac{100}{\gamma}$ is chosen as before.
We use a viscosity of $\mu = 0.01$ and analytical boundary conditions.
Our results (\cref{fig:abc-flow}) show a good agreement between the analytical solution and our approximation with an \aderdg{}-scheme of order $3$ with a mesh consisting of $27^3$ cells at time $t = \SI{0.1}{\s}$.
Again, we do not perform a quantative analysis as the \textsc{abc}-flow only solves our equation set approximately.

@@

\subsection{Stratified Flow Scenarios}
With the constants
\begin{equation}\label{eq:atmosphere-constants}
    \gamma = 1.4 ,\quad \Pr =  0.71 ,\quad R = 287.058 ,\quad p_0 = \SI{10e5}{\Pa}, \quad g = \SI{9.8}{m/s^2},
\end{equation}
all following stratified flow scenarios are described in terms of the potential temperature
\begin{equation}
  \potT = T \left( \frac{p_0}{p} \right)^{R/c_p},
\end{equation}
with reference pressure $p_0$~\cite{muller2010adaptive,giraldo2008study}.
%
We compute the initial background density and pressure by inserting the assumption of a constant background energy in \cref{eq:hydrostatic-balance}.
The background atmosphere is then perturbed.
We set the density and energy at the boundary such that it corresponds to the background atmosphere.
Furthermore, to ensure that the atmosphere stays in hydrostatic balance, we need to impose the viscous heat flux
\begin{equation}
  \label{eq:atmosphere-bc}
  \viscFlux_{\QE} = \kappa \pdv{\overline{T}}{z}.% =
\end{equation}
at the boundary~\cite{giraldo2008study}.
In this equation, $\overline{T}(z)$ is the background temperature at position $z$, which can be computed from \cref{eq:hydrostatic-balance,eq:eos}.

@@

We have two bubbles, with constants
\begin{equation}
  \label{eq:bubbles-values}
\begin{alignedat}{6}
  & \text{warm:} \qquad && A = \SI{0.5}{\K}, \quad&& a = \SI{150}{\m}, \quad&& s = \SI{50}{\m}, \quad&& x_c = \SI{500}{\m,} \quad&& z_c = \SI{300}{\m},\\
  & \text{cold:} \qquad && A = \SI{-0.15}{\K}, \quad&& a = \SI{0}{\m}, \quad&& s = \SI{50}{\m}, \quad&& x_c = \SI{560}{\m}, \quad&& z_c = \SI{640}{\m}.
  \end{alignedat}
\end{equation}
Similar to~\cite{muller2010adaptive}, we use a constant viscosity of $\mu = 0.01$ to regularize the solution.
Note that we use a different implementation of viscosity than~\cite{muller2010adaptive}.
Hence, it is difficult to compare the parametrization directly.
We ran this scenario twice:
once without \amr{} and a mesh of size $\SI{1000/81}{\m} = \SI{12.35}{\m}$ and once with \amr{} with two adaptive refinement levels and parameters $\Trefine = 2.5$ and $\Tdelete = -0.5$.
For both settings we used polynomials of order 6.
We specialize the \amr{}-criterion~\pcref{eq:refinement-criterion} to our stratified flows by using the potential temperature.
This resulted in a mesh with cell-size lengths of approx.\ \SI{111.1}{\m}, \SI{37.04}{\m}, and \SI{12.34}{\m}.
The resulting mesh can be seen in \cref{fig:two-bubbles-ader}.
We observe that the $L_2$ difference between the potential temperature of the \amr{} run, which uses 1953 cells, and the one of the fully refined run with 6561 cells, is only $1.87$.
The relative error is \num{6.69e-6}\todo{Update these numbers}.
We further emphasize that our \amr{}-criterion accurately tracks the position of the edges of the cloud instead of only its position.
This is the main advantage of our gradient-based method in contrast to methods working directly with the value of the solution, as for example~\cite{muller2010adaptive}.
Overall, our result for this benchmark shows an excellent agreement to the previous solutions of~\cite{muller2010adaptive}.

@@

For our second scenario, the cosine bubble, we use a perturbation of the form
\begin{align}
  \label{eq:cos-pertubation}
  \pertubationPotT &= \begin{cases}
    \nicefrac{A}{2} \left[ 1 + \cos(\pi r) \right] & r \leq a, \\
    0 & r > a,
    \end{cases}
\end{align}
where $A$ denotes the maximal perturbation and $a$ is the size of the bubble.
We use the constants
\begin{equation}\label{eq:cosine-bubble}
  A = \SI{0.5}{\K} \quad a = \SI{250}{\m} \quad x_c = \SI{500}{\m} \quad z_c = \SI{350}{\m}.
\end{equation}
For the three-dimensional bubble, we set $y_c = x_c = \SI{500}{\m}$.
This corresponds to the parameters used in~\cite{kelly2012continuous}\footnote{%
We found that the parameters presented in the manuscript of~\cite{kelly2012continuous} only agree with the results, if we use the same parameters as for 2D simulations.}.
For the 2D case, we use a constant viscosity of $\mu = 0.01$ and an \aderdg{}-method of order 6 with two levels of dynamic \amr{}, resulting again in cell sizes of roughly $\SI{111.1}{\m}, \SI{37.04}{\m}, \SI{12.34}{\m}$.
We use slightly different \amr{} parameters of $\Trefine = 1.5$ and $\Tdelete = -0.5$ and let the simulation run for \SI{600}{\s}.
Note that, as seen in \cref{fig:cosine-2d}, our \amr{}-criterion tracks the wavefront of the cloud accurately.
This result shows an excellent agreement to the ones achieved in~\cite{giraldo2008study,muller2010adaptive}.

@@

\subsection{Performance Results}
\todo{1--2 days on workstation}
\todo{tbb scalability in MDOF/s}
\todo{Resolution low enough: weak scaling would be more interesting}
All two-dimensional scenarios presented in this paper can be run on a single workstation in less than two days.
Parallel scalability was thus not the primary goal of this paper.
Nevertheless, our implementation allows us to scale to small to medium scale setups using a combined MPI/TBB parallelization strategy.

\newcommand{\mdofs}{M\text{DOF}/s}%
We typically first choose a number of MPI ranks that ensure an equal load balancing.
For two-dimensional simulations this happens for $1, 10, 83, \ldots$ ranks, as our underlying framework uses a three-way splitting per dimension and an additional communication rank per level.
Then, we choose the number of nodes that we want to use and divide the cores into shared-memory sections.
These are then distributed among TBB ranks.
We ran the two bubble scenario for a uniform grid for 30 timesteps and for multiple combinations of MPI ranks and TBB threads.
This simulation was performed on the SuperMUC-NG system using computational kernels that are optimized for its Skylake architecture\todo{Cite opt. kernels?}.
Using a single MPI rank, we have xxx x $10^x$ degrees of freedom updates ($\mdofs$) using two TBB threads and xxx $10^x \mdofs$ using xxx threads.
When using 5 nodes, we achieve $x \mdofs$ for two threads and $x \mdofs$ for xxx threads, when using two ranks per node.

@@

Future work should be directed towards improving the scalability.
As mentioned in the results section, the resolution is sufficient for most numerical weather prediction purposed.
With an improved \amr{} scaling and some fine tuning of the parallelization strategy, the numerical method presented here might be a good candidate for the small-scale simulation of clouds.
